---
output-file: case-study-three.html
lightbox: auto
---

```{r}
#| label: setup
#| echo: false

suppressPackageStartupMessages(source(here::here("packages.R")))
for (f in list.files(here::here("R"), full.names = TRUE)) source (f)
```

# The Use of Data in Local Governance - A Michigan Perspective {#sec-cs3-michigan}

This case study explores the current state of data-driven decision-making in Michigan's local governments, highlighting challenges and opportunities for integrating data into policy and governance based on the results of the [Michigan Public Policy Survey (MPPS)](https://closup.umich.edu/michigan-public-policy-survey)\index{Michigan Public Performance Survey}[@ivacko_data-driven_2013].

## Michigan Public Policy Survey {#sec-mpps}

The MPPS, established post the 2009 Great Recession, is the first ongoing survey of local leaders across an entire state in the United States, involving over 1,856 jurisdictions in Michigan. It addresses a critical gap by providing insights into local officials' perspectives, crucial for informed policymaking. Conducted biannually, it tracks long-term trends on fiscal and operational policies while addressing current issues like the COVID-19 pandemic and infrastructure. Collaborations with key associations enhance its credibility and scope.

## Current Situation of Policy and Decision-Making in Michigan Local Governments {#sec-cs3-current-situation}

Michigan's local governments have seen significant growth in data-driven decision-making (see @fig-mpss-01 and @fig-mpss-02). 

```{r}
#| label: fig-mpss-01
#| fig-cap: "Percentage of Michigan jurisdictions reporting use of performance data"
#| fig-height: 6
#| column: margin
#| echo: false

mpss <- read.csv("data/f11data-use_pud.csv")

mpss |>
  count(q2_new) |>
  mutate(n_total = sum(n), p = (n / n_total) * 100) |>
  mutate(
    q2_new = case_when(
      q2_new == 1 ~ "Use data about\nown or other\njurisdictions",
      q2_new == 2 ~ "Do not use data\nabout own or\nother jurisdictions",
      q2_new == 3 ~ "Don't know"
    ) |>
      factor(
        levels = c(
          "Use data about\nown or other\njurisdictions",
          "Do not use data\nabout own or\nother jurisdictions",
          "Don't know"
        )
      )
  ) |>
  ggplot(mapping = aes(x = p, y = q2_new, fill = q2_new)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_x_continuous(limits = c(0, 100)) +
  scale_fill_manual(
    name = NULL, values = oxford_theme_palettes()$contemporary[2:4]
  ) +
  labs(x = "%", y = NULL) +
  theme_oxford(
    base_family = "", grid = "X", grid_col = get_oxford_colour("stone")
  )
```

```{r}
#| label: fig-mpss-02
#| fig-cap: "Michigan jurisdictions’ history of data use (among current data users), by population density"
#| fig-width: 4
#| column: margin
#| echo: false

mpss |>
  count(pop_density, q4) |>
  filter(!is.na(q4)) |>
  group_by(pop_density) |>
  mutate(n_total = sum(n), p = (n / n_total) * 100) |>
  mutate(
    q4 = case_when(
      q4 == 1 ~ "Less than\n1 year",
      q4 == 2 ~ "Between\n1-5 years",
      q4 == 3 ~ "Longer than\n5 years",
      q4 == 4 ~ "Don't know"
    ) |>
      factor(
        levels = c(
          "Less than\n1 year",
          "Between\n1-5 years",
          "Longer than\n5 years",
          "Don't know"
        )
      ),
    pop_density = case_when(
      pop_density == 1 ~ "Low",
      pop_density == 2 ~ "Medium",
      pop_density == 3 ~ "High"
    ) |>
      factor(levels = c("Low", "Medium", "High"))
  ) |>
  ggplot(mapping = aes(x = pop_density, y = p, fill = q4)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    name = NULL, values = oxford_theme_palettes()$celebratory[1:4]
  ) +
  labs(x = "Population Density", y = "%") +
  theme_oxford(
    base_family = "", grid = "Y", grid_col = get_oxford_colour("stone")
  ) +
  theme(legend.position = "top", legend.text = element_text(size = 8))
```

This approach is now widespread across jurisdictions of all population sizes (see @fig-mpss-03) and across regions, with many jurisdictions using data to inform budgeting and resource allocation. 

```{r}
#| label: fig-mpss-03
#| fig-cap: "Percentage of Michigan jurisdictions reporting data use, by population density"
#| fig-width: 4
#| column: margin
#| echo: false

mpss |>
  count(pop_density, q2_new) |>
  group_by(pop_density) |>
  mutate(n_total = sum(n), p = (n / n_total) * 100) |>
  mutate(
    q2_new = case_when(
      q2_new == 1 ~ "Use data about\nown or other\njurisdictions",
      q2_new == 2 ~ "Do not use data\nabout own or\nother jurisdictions",
      q2_new == 3 ~ "Don't know"
    ) |>
      factor(
        levels = c(
          "Use data about\nown or other\njurisdictions",
          "Do not use data\nabout own or\nother jurisdictions",
          "Don't know"
        )
      ),
    pop_density = case_when(
      pop_density == 1 ~ "Low",
      pop_density == 2 ~ "Medium",
      pop_density == 3 ~ "High"
    ) |>
      factor(levels = c("Low", "Medium", "High"))
  ) |>  
  ggplot(mapping = aes(x = pop_density, y = p, fill = q2_new)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    name = NULL, values = oxford_theme_palettes()$contemporary[2:4]
  ) +
  labs(x = "Population Density", y = "%") +
  theme_oxford(
    base_family = "", grid = "Y", grid_col = get_oxford_colour("stone")
  ) +
  theme(legend.position = "top", legend.text = element_text(size = 8))
```

Despite this progress, most data use remains informal or ad hoc (see @fig-mpss-04), particularly among smaller communities (see @fig-mpss-05).

```{r}
#| label: fig-mpss-04
#| fig-cap: "Percentage of Michigan jurisdictions reporting ad hoc vs. systematic data use (among data users)"
#| fig-height: 6
#| column: margin
#| echo: false

mpss |>
  count(q3) |>
  filter(!is.na(q3)) |>
  mutate(n_total = sum(n), p = (n / n_total) * 100) |>
  mutate(
    q3 = case_when(
      q3 == 1 ~ "Use data on ad hoc basis",
      q3 == 2 ~ "Use data as part of a\nformal program for\nsome or all operations",
      q3 == 3 ~ "Don't know"
    ) |>
      factor(
        levels = c(
          "Use data on ad hoc basis",
          "Use data as part of a\nformal program for\nsome or all operations",
          "Don't know"
        )
      )
  ) |>
  ggplot(mapping = aes(x = p, y = q3, fill = q3)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_x_continuous(limits = c(0, 100)) +
  scale_fill_manual(
    name = NULL, values = oxford_theme_palettes()$heritage[3:5]
  ) +
  labs(x = "%", y = NULL) +
  theme_oxford(
    base_family = "", grid = "X", grid_col = get_oxford_colour("stone")
  )
```

```{r}
#| label: fig-mpss-05
#| fig-cap: "Percentage of Michigan jurisdictions reporting ad hoc vs. systematic data use (among data users), by population size"
#| fig-width: 4
#| column: margin
#| echo: false

mpss |>
  count(pop_density, q3) |>
  filter(!is.na(q3)) |>
  group_by(pop_density) |>
  mutate(n_total = sum(n), p = (n / n_total) * 100) |>
  mutate(
    q3 = case_when(
      q3 == 1 ~ "Use data on ad hoc basis",
      q3 == 2 ~ "Use data as part of a\nformal program for\nsome or all operations",
      q3 == 3 ~ "Don't know"
    ) |>
      factor(
        levels = c(
          "Use data on ad hoc basis",
          "Use data as part of a\nformal program for\nsome or all operations",
          "Don't know"
        )
      ),
    pop_density = case_when(
      pop_density == 1 ~ "Low",
      pop_density == 2 ~ "Medium",
      pop_density == 3 ~ "High"
    ) |>
      factor(levels = c("Low", "Medium", "High"))
  ) |>
  ggplot(mapping = aes(x = pop_density, y = p, fill = q3)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    name = NULL, values = oxford_theme_palettes()$heritage[3:5]
  ) +
  labs(x = "Population Density", y = "%") +
  theme_oxford(
    base_family = "", grid = "Y", grid_col = get_oxford_colour("stone")
  ) +
  theme(legend.position = "top", legend.text = element_text(size = 8))
```

The MPPS reveals that while larger jurisdictions are more likely to engage in formal performance measurement, over half of the state's smallest jurisdictions also incorporate some form of data into their decision-making processes (see @fig-mpss-02). This indicates a trend towards broader adoption, albeit at varying levels of formality.

```{r}
#| label: fig-mpss-06
#| fig-cap: "Michigan jurisdictions’ plans for expanding future data use (among current data users), by population density"
#| fig-cap-location: top
#| fig-width: 4
#| echo: false
#| eval: false

mpss |>
  count(pop_density, q14) |>
  filter(!is.na(q14)) |>
  group_by(pop_density) |>
  mutate(n_total = sum(n), p = (n / n_total) * 100) |>
  filter(q14 %in% 4:5) |>
  mutate(
    q14 = case_when(
      q14 == 4 ~ "Likely to expand\nsomewhat",
      q14 == 5 ~ "Likely to expand\nsignificantly"
    ) |>
      factor(
        levels = c(
          "Likely to expand\nsomewhat", "Likely to expand\nsignificantly"
        )
      ),
    pop_density = case_when(
      pop_density == 1 ~ "Low",
      pop_density == 2 ~ "Medium",
      pop_density == 3 ~ "High"
    ) |>
      factor(levels = c("Low", "Medium", "High"))
  ) |>
  ggplot(mapping = aes(x = pop_density, y = p, fill = q14)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    name = NULL, values = oxford_theme_palettes()$corporate[3:4]
  ) +
  labs(x = "Population Density", y = "%") +
  theme_oxford(
    base_family = "", grid = "Y", grid_col = get_oxford_colour("stone")
  ) +
  theme(legend.position = "top", legend.text = element_text(size = 8))
```

## Challenges and Concerns {#sec-cs3-challenges}

1. **Cost Concerns**

Many local governments, especially smaller ones with limited resources, perceive data use as costly. The MPPS found that 62% of non-data users cited cost concerns, though only 28% of current users reported significant issues, suggesting costs may be manageable.

2. **Informal Practices**

The reliance on informal methods can lead to inconsistent outcomes and less accountability. Only about 16% of jurisdictions have formal performance measurement practices, indicating a gap in structured data use.

3. **Resource Constraints**

Smaller jurisdictions often face limitations in staff and financial resources, hindering their ability to adopt more formal data practices.

## Opportunities and Benefits {#sec-cs3-opportunities}

1. **Fiscal Efficiency**

Data-driven approaches help identify cost savings and program efficiencies, crucial for jurisdictions grappling with fiscal challenges.

2. **Improved Service Delivery**

By aligning services with community needs, data can enhance service quality and responsiveness.

3. **Enhanced Transparency and Trust**

Effective use of data fosters transparency, improving public trust in government decisions.

4. **Policy Communication**

Data provides a clear evidence base for policy-making, aiding communication between governments and stakeholders.

## Conclusion {#sec-cs3-conclusion}

The integration of data into Michigan's local governance has proven valuable despite challenges like cost concerns and resource limitations. The broader adoption of data-driven practices, even informally, highlights its potential to improve decision-making and service delivery.

## Recommendations {#sec-csc3-recommendations}

1. **Capacity Building** - Invest in training to enhance technical and analytical skills among local officials.

2. **Encourage Collaboration** - Foster partnerships with academic institutions or tech firms to support data initiatives.

3. **Leverage Resources** - Utilise available tools and frameworks, such as those provided by Michigan's MPPS, to guide data practices.

4. **Promote Leadership and Cultural Change** - Champion leadership roles that prioritize data use and cultivate a culture of evidence-based decision-making.

By adopting these strategies, countries can effectively integrate data into local governance, enhancing policy outcomes and public trust.

<br/>
<br/>